Supplementary Materials to ``Recurrent event analysis in the presence of real-time high frequency data via random subsampling''

### 2023-02-02 letters/response-ref-1-v1.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Plain Cover Letter
% LaTeX Template
%
% This template has been downloaded from:
% http://www.latextemplates.com
%
% Original author:
% Rensselaer Polytechnic Institute (http://www.rpi.edu/dept/arc/training/latex/resumes/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[11pt]{letter} % Default font size of the document, change to 10pt to fit more text

\usepackage{newcent} % Default font is the New Century Schoolbook PostScript font
%\usepackage{helvet} % Uncomment this (while commenting the above line) to use the Helvetica font

% Margins
\topmargin=-1in % Moves the top of the document 1 inch above the default
\textheight=8.5in % Total height of the text on the page before text goes on to the next page, this can be increased in a longer letter
\oddsidemargin=-10pt % Position of the left margin, can be negative or positive if you want more or less room
\textwidth=6.5in % Total width of the text, increase this if the left margin was decreased and vice-versa

\let\raggedleft\raggedright % Pushes the date (at the top) to the left, comment this line to have the date on the right
\usepackage{amsthm, amsfonts}
\usepackage{xcolor}
\usepackage{hyperref}

\def\Y{{\bf Y}}

\begin{document}

%----------------------------------------------------------------------------------------
%	ADDRESSEE SECTION
%----------------------------------------------------------------------------------------

\begin{letter}{Professor
	Peter Glynn\\
	Editor, {\em Journal of Applied Probability}}

%----------------------------------------------------------------------------------------
%	YOUR NAME & ADDRESS SECTION
%----------------------------------------------------------------------------------------

%\begin{center}
%\large
%\end{center}
%\vfill

\signature{Walter Dempsey\\
University of Michigan\\
Department of Biostatistics\\
1415 Washington Heights\\
Ann Arbor, MI 48103} % Your name for the signature at the bottom

%----------------------------------------------------------------------------------------
%	LETTER CONTENT SECTION
%----------------------------------------------------------------------------------------

\vspace{5mm}

\newpage

{\bf Response to Referees}

We appreciate all of the very helpful comments from the editor, associate editor, and the two referees which has helped improve the paper substantially. We have addressed the concerns as best as possible in this revision. We provide a further point-by-point explanation below. Any comments in {\it italics} indicate a statement from the referee report.

{\bf Response to Associate Editor}
\begin{enumerate}
\item {\it This paper was reviewed by two referees, who tend to agree in their positive assessments of the value and novelty of your work, and provide some very specific comments for further improvements.}

\vspace{5mm}
We thank the associate editor.  We have addressed the two reviewers' specific comments.  Responses to their concerns are detailed below and have significantly improved the clarify of this manuscript.
\vspace{5mm}

\item {\it Expanded literature review on existing modeling approaches. Both referees raise this point, and I agree that the contrast with existing works could be more explicit. The second referee provides some specific suggestions on relevant works.}

\vspace{5mm}
XXXX
\vspace{5mm}

\item {\it There appear to be some missing definitions of notation, and some transitions/equations need more details to justify them. The latter could potentially be put in the supplement to save space. The former should be added to the main manuscript to improve clarity.}

\vspace{5mm}
XXXX
\vspace{5mm}

\item {\it As the first reviewer, I am confused by $\varepsilon(t) = 0$ on page 4, line 39, in line with equation (3) and measurement error methodology. I suspect comment 7 from the 2nd reviewer alludes to the same thing. I suspect this may be a notation issue rather than the modeling issue, but either way, it needs to be clarified.}

\vspace{5mm}
XXXX
\vspace{5mm}

\item {\it Practical considerations, such as window size and smoothing parameter selection, need clarification.}

\vspace{5mm}
XXXX
\vspace{5mm}

\item {\it Reviewer 2 suggests adding additional details on proof steps and lemma clarifications. I can't entirely agree that the proofs should go in the main manuscript as the paper is already on the longer side. When making adjustments to this section (theoretical analysis), please keep at most its current length, and move things to supplement as needed.}

\vspace{5mm}
XXXX
\vspace{5mm}

\item {\it Both reviewers suggest additional simulation studies. I agree that it would be good to have a numerical benchmark of the magnitude of the computational issue that you allude to in lines 16-28 on page 5, especially for JCGS readership. I don’t think you need references for this, as the reviewer asks, as a numerical example would be more helpful and convincing in this case. However, I think you need another method for comparison, as the simulations are only done for your approach with different sampling rates. Perhaps some suggestions from the first referee could be useful. If some of these alternatives are computationally prohibitive, then doing a computational numerical comparison on even 1 rep will provide a convincing argument for why those methods are not tried.}

\vspace{5mm}
XXXX
\vspace{5mm}

\item {\it I also have additional comments that were not raised. A major one is the absence of code. A user-friendly software with the method’s implementation and documentation is a must for accepted JCGS papers. Please provide the code so that the results are reproducible.}

\vspace{5mm}
XXXX
\vspace{5mm}

\item {\it
A minor one is the absence of truncation $K_x$ in the equation on line 37, page 8. Please reread everything carefully to ensure all notations are defined and the equations are consistent with the text.}

\vspace{5mm}
XXXX
\vspace{5mm}

\item {\it Another minor one is your choice of smoothing the entire covariance, including the diagonals (page 8, lines 13-18). It’s unclear to me what the rationale is for changing this step compared to Park and Staicu, 2015, and how much it matters for the final performance.}

\vspace{5mm}
XXXX
\vspace{5mm}

\item {\it Expanded literature review on existing modeling approaches. Both referees raise this point, and I agree that the contrast with existing works could be more explicit. The second referee provides some specific suggestions on relevant works.}

\vspace{5mm}
XXXX
\vspace{5mm}

\end{enumerate}

\newpage
{\bf Response to Reviewer 1}
\vspace{5mm}

I am very grateful to the referee for the incredibly detailed review.  The detailed suggestions and comments helped improve the paper in numerous ways.  See below for our responses to each particular reviewer comment.

\begin{enumerate}
\item {\it The introduction it quite lacking. Please add more background literature review of prevailing/latest approaches (particularly in the area of joint modeling of longitudinal and time to event data) to the introduction.}

\vspace{5mm}
XXXXX
\vspace{5mm}

\item {\it The measurement error part can be added to the end with other exten-
sions. It does not seem to play a major role in developing the algorithm.}

\vspace{5mm}
We agree.  This point was over-emphasized in the original manuscript and so has been moved to the
\vspace{5mm}

\item {\it In equationn (4), page 4, do you mean that equation (2) and the mesurement error part $x_i (t) = \eta_i (t) + \epsilon_i (t)$ imply conditional hazard equation in (3)?  If yes, then why is $\int_{t-\Delta}^t \epsilon_i (s) \beta(s) ds = 0$?  Or are you assuming a model?  Please add more clarity.}

\vspace{5mm}
XXXXX
\vspace{5mm}

\item {\it Several notations in section 2.1 such as ${\bf x}_i, {\bf H}_{i,\tau_i}^X$ are not defined.}

\vspace{5mm}
XXXXX
\vspace{5mm}

\item {\it How did you get the equation in line 24, page 4.}

\vspace{5mm}
XXXXX
\vspace{5mm}

\item {\it In line 39, page 4, by assuming $\epsilon(t) = 0$, you assume there is no measure- ment error? Why is section 2.1 called measurement error if the error is 0? Should this be named missing data section instead?}

\vspace{5mm}
The key point raised on page 4 (no Section XX), is that if $\epsilon(t)$ is not equivalently zero, your model assumes risk of an event at time $t$ depends on both the current and future measurements of the process ${\bf x}_i$. We argue this is not sensible since current risk should not depend on future risk and thus set $\epsilon(t) = 0$.
\vspace{5mm}

\item  {\it Is $\theta$ defined anywhere prior to section 2.2?}

\vspace{5mm}
It was used in equation (1) to refer to all the parameters underlying the model to distinguish it from the functional parameter~$\beta (s)$.
We do XXXX to make this clearer.
\vspace{5mm}

\item {\it On page 5, lines 16 - 28 you claim that computing the double integral is computationally expensive. Add some references. Also, to give a greater insight to the extent of the computational cost, solve the score equation in lines 8,9 i.e. without the sub-sampling framework in the simulations (there is another comment on this for the simulation section). Reporting a rough summary of computational cost such as ” it took x min to solve equations involving n samples” here, would help the reader understand
the extent of the problem.}

\vspace{5mm}
We explain that if we were to compute it would have to be as a function of $\beta(s)$.  We implement a naive and demonstrate numerically the computational cost.  Specifically, each computation in simulation takes XXseconds, and we must compute this XX times to compute the likel
\vspace{5mm}

\item {\it Page 5, line 51, what is $t_i$?}

\vspace{5mm}
This should be $\tau_i$ which is the censoring time of the observation process for individual~$i$.
\vspace{5mm}

\item {\it Check the notation in equation (6) page 6. Some subscripts seem to be missing. Add some details on how to obtain this from the original score
equation. What is the approximation here?}

\vspace{5mm}
We have included correct subscripts in the revision.  Derivation of (6) follows from plugging in (4) into equation the score equations.  We now make this clear and provide a derivation in the supplementary materials for completeness.
\vspace{5mm}

\item{\it The remark 2.1 can be added the end of section 2.2. This along with
comment 8 will naturally lead the way your proposed method.}

\vspace{5mm}
We thank the referee for this suggestion.  We have now used Remark 2.1 and comment 8 on the computational costs to the end of section 2.2 to motivated our proposed method.
\vspace{5mm}

\item {\it Page 7, line 48, in the definition of the marginal covariance, it should be $dT$ in the integral.}

\vspace{5mm}
That is correct.  We have fixed the definition as suggested.
\vspace{5mm}

\item {\it Page 7, line 48, in the definition of the marginal covariance, what is $\tau$?}

\vspace{5mm}
The term $\tau$ refers to the censoring time of the observation process. We make this clear in Section XXX, page XX in the revision.
\vspace{5mm}

\item {\it s there a difference between the mean and covariance functions for $y = 0$ and $y = 1$ especially if points in $D_i$ and $T_i$ are spread roughly evenly on a grid even if the sets are disjoint? For example, pick a grid of equidistant points and number them from $1-10$. If all odd points are events in $T_i$ and even ones are sampled in $D_i$, won't the pooled sample covariance for both be the same}

\vspace{5mm}
The event times~$T_i$ depend heavily on the event intensity function~$h_i (t;\theta)$ and will not be guaranteed to be uniformly spread across points in $[0,\tau]$.  The subsampling times~$D_i$ depend on the subsampling intensity function~$\pi_i(t)$ and
\vspace{5mm}

\item {\it Add a plot of the mean, covariance functions for $y=0$ and $y=1$.}

\vspace{5mm}
We have added mean and covariance functions to the supplementary materials as suggested.  The plots demonstrate distinct behavior in the means for $y=0$ and $y=1$; while the covariance functions are XXXX.
\vspace{5mm}

\item {\it Page 9, line 6, isn’t the integral in (2) $\int_{t-\Delta}^t X(s) \beta(s)ds$? The whole development assumes the integral to be $\int_{t-\Delta}^t X(t−s) \beta(s)ds$. The interpretation of the coefficient function $\beta()$ would be different for two cases. Please clarify.}

\vspace{5mm}
We apologize for the confusion.  In Section 3.1, we started by defining the double-indexed $X(t,s)$ to be equal to $X(t-s)$ in order to make the notation a bit cleaner but we see that this has added confusion.  We not add a remark to make this more clear.  The reason for the notation was that writing $X(t,s)$ allows us to write the integrals over $0$ to $\Delta$ rather than having to constantly index integrals by $t$.  Moreover, we thought it helps the reader to understand we are saying ``the process $s$ units prior to time~$t$.''
\vspace{5mm}

\item {\it Page 9, equation 7, please define all of the notations before using them.}

\vspace{5mm}
We have re-written this paragraph to first define the notation and then present equation (7).
\vspace{5mm}

\item {\it Page 12, line 48, what is proposition 3.6? You mean Lemma?}

\vspace{5mm}
That is correct, we have fixed this issue.
\vspace{5mm}

\item {\it When are the assumptions satisfied or not satisfied? An example of a
situation where the functional and event process assumptions are easily satisfied will be useful.}

\vspace{5mm}
We have shown that the functional process satisfies these moment conditions under boundedenss which is
\vspace{5mm}

\item {\it All the modes of convergences in Lemma 3.5 and its proofs should be made clear.}

\vspace{5mm}
XXXxXX
\vspace{5mm}

\item {\it Outline the proof of Lemma 3.5 first and then prove the details.  It will be easier to follow.  Alternatively, we }

\vspace{5mm}
We have added a sketch of the proof with an outline as part of the supplementary details.
\vspace{5mm}

\item {\it In Lemma 3.6 state the theorem clearly.  What condition do yu show in the proof that establishes optimality}

\vspace{5mm}
We have now included the statement of how we define optimality in the main paper to make the theorem statement clear.  Proof is kept in the supplementary materials for conciseness.
\vspace{5mm}

\item {\it Combine section 3.6 and 3.7 in to a single result that establishes a confidence band for $\beta (t)$. Also, wont Lemma 3.5 lead to a confidence band for $\hat \beta (t)$? Are the two bands same?}

\vspace{5mm}
We discuss the penalization due to the potential complexity of the.  The confidence bands are based on Lemma 3.5.  We make this clear in the revision.  The issues are still thre and we keep the discussion.  We keep as 2 separate sections since they cover different issues.
\vspace{5mm}

\item The ``original'' score equation Page 5, line 9 involves
$$
H_i (\tau_i ;\theta) = \int_{t=0}^{\tau_i} h_i (t; \theta)dt = \int_0^{\tau_i} \left[ h_0 (t;\gamma) \exp \left( g_t (H_{i,t}^N)^\top \alpha + \int_{s=t-\Delta}^t x_i (s) \beta(t) ds \right) \right] dt
$$
In the subsampled version you suggest in equation 6, page 6, you essentially replace $H_i (\tau_i; \theta)$ computation with $h_i (u; \theta)$, $u \in D_i$. How much time would it take to solve the full likelihood equation that will involve the above equation (1) directly? With the basis expansion strategy, you
would will be estimating the same number of parameters. This type of approach should be added to simulations and evaluated for accuracy and time complexity.

\vspace{5mm}
XXXXX
\vspace{5mm}

\item {\it How about using a concurrent model for $h_i (t;\theta) = h_0 (t; \theta) \exp (g_t (H_{i,t}^N)^\top \alpha + x_i (t) \beta (t))$. This does not look at the past window like the suggested model but this drawback might be compensated with the constructed features $g_t (H_{i,t}^N)$ that capture the past. The concurrent approach might be much less expensive computationally. This can be added as an alternative to the simulations.}

\vspace{5mm}
XXXXX
There is no obvious way to compute bias since the model will be incorrectly specified.  We do show the computation trade-off and that computing the cumulative hazard is still computationally demanding; therefore, after sampling non-event times, the added computation of the interior integral is not very expensive and avoids model misspecification.
\vspace{5mm}

\item {\it In section 5.2.1 what is uncongeniality? Does it have a mathematical
meaning?}

\vspace{5mm}
uncongenial comes from Meng. We cite and provide a short formal definition with some intuition.
\vspace{5mm}

\item Page 23, Figure 2. EDA seems to spike at the end. Even AI seems to be
turning up. Add longer history to both the Figures.

\item {\it Doesn’t Figure 3 suggest smaller ∆?}

\vspace{5mm}
We agree.  We re-run the analysis with a window length of 15 minutes as well and move the current analysis as a sensitivity to the supplementary materials.
\vspace{5mm}

\item {\it What features were constructed for the real data analysis?}

\vspace{5mm}
As this was illustrative we only included time-of-day.
\vspace{5mm}

\end{enumerate}

\newpage

{\bf Response to Reviewer \#2}

I am incredibly grateful for this review.  I agree that the paper has evolved beyond the current title.  This was not updated in the prior submission simply to ensure continuity of presentation (i.e., didn't want to change from first submission).  Here we replace with the title ``Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information''.  We next address the reviewers specific comments.

\begin{enumerate}
\item {\it The paper states some facts about the SARS-CoV-2 virus that are outdated, for example “Upon infection with SARS-CoV-2, an incubation period starts and lasts approximately five days [Lauer et al., 2020] before the viral load is high enough to be detected.” (Page 5). Lauer was based on the original
strain, we see different disease dynamics with the delta strain. Perhaps to be more precise, the author could clarify that this is an example of the original SARS-CoV-2 incubation rather than definitative as there will likely be future variants that will change this further. Similarly, on page 13, the serial
interval of 7 is stated as fact – this is again based on the original strain rather than the current disease dynamics. Stating these numbers as an example rather than definitive would be preferable, I believe.}
\vspace{5mm}
\end{enumerate}

\end{letter}

\end{document}

### 2023-02-02 write-up/fda-recurrent-supplement.tex
updating paper first, then code for revision

@@

\title[Recurrent event analysis with functional covariates via random subsampling]{Recurrent event analysis in the presence of real-time high frequency data via random subsampling}
% \author{Walter Dempsey}
% \address {Department of Biostatistics, University of Michigan, 1415 Washington Heights, Ann Arbor, MI 48109, USA}
% \email{wdem@umich.edu}

@@

\begin{abstract}
Digital monitoring studies collect real-time high frequency data via mobile sensors in the subjects' natural environment. This data can be used to model the impact of changes in physiology on recurrent event outcomes such as smoking, drug use, alcohol use, or self-identified moments of suicide ideation. Likelihood calculations for the recurrent event analysis, however, become computationally prohibitive in this setting. Motivated by this, a random subsampling framework is proposed for computationally efficient, approximate likelihood-based estimation. A subsampling-unbiased estimator for the derivative of the cumulative hazard enters into an approximation of log-likelihood. The estimator has two sources of variation: the first due to the recurrent event model and the second due to subsampling. The latter can be reduced by increasing the sampling rate; however, this leads to increased computational costs.  The approximate score equations are equivalent to logistic regression score equations, allowing for standard, ``off-the-shelf'' software to be used in fitting these models. Simulations demonstrate the method and efficiency-computation trade-off. We end by illustrating our approach using data from a digital monitoring study of suicidal ideation.
\end{abstract}

@@

\section{Introduction}

Advancement in mobile technology has led to the rapid integration of mobile and wearable sensors into behavioral health~\citep{Freeetal2013}. Take HeartSteps, for example, a mobile health (mHealth) study designed to increase physical activity in sedentary adults \citep{KlasnjaHS2019}. Here, a Jawbone sensor is used to monitor step count every minute of the participant's study day. Of interest in many mHealth studies is the relation of such real-time high frequency sensor data to an adverse, recurrent event process. In a smoking cessation mHealth study~\citep{Sense2Stop}, for example, the relation between a time-varying sensor-based measure of physiological stress and smoking lapse is of scientific interest. In a suicidal ideation mHealth study~\citep{Kleiman2018}, the relation of electrodermal activity (EDA) and accelerometer with self-identified moments of suicidal ideation is of scientific interest.

The goal of this paper is two-fold: (1) to discuss the appropriate choice of statistical model for joint high-frequency sensor data and the recurrent event data, and (2) to construct a simple, easy-to-implement method for parameter estimation and inference. For (1), we discuss an important issue regarding measurement-error models when paired with recurrent event outcomes. For (2), we introduce a random subsampling procedure that has several benefits.  First, the resulting inference is unbiased; however, there is a computation-efficiency trade-off. In particular, a higher sampling rate can decrease estimator variance at the cost of increased computation.  We show via simulations that the benefits of very high sampling rates is often negligible, as the contribution to the variation is small relative to the variation in the underlying stochastic processes. Second, derived estimating equations are optimal, implying loss of statistical efficiency is only due to subsampling procedure and not the derived methodology.  Finally, implementation can leverage existing, standard software for functional data analysis and logistic regression, leading to fast adoption by domain scientists.

\section{Recurrent event process and associated high frequency data}

Suppose $n$ subjects are independently sampled with observed event times~$\bfT_{i} = \{ T_{i,1}, \ldots, T_{i,k_i}\}$ over some observation window $[0, \tau_i]$ for each subject $i = 1,\ldots, n$.  Assume the event times are ordered, i.e., $T_{i,j} < T_{i,j^\prime}$ for $j < j^\prime$. The observation window length, $\tau_i$, is the censoring time and is assumed independent of the event process. Let~$N_{i} (t)$ denote the associated counting process of $\bfT_{i}$; that is, $N_i (t) = \sum_{j=1}^{k_i} 1 [ T_{i,j} < t ]$. In this section, we assume a single-dimensional health process~$\bfx_i = \{ x_i (s) \}_{0 < s < \tau_i}$ for each participant is measured at a dense grid of time points.  Accelerometer, for example, is measured at a rate of 32Hz (i.e., $32$ times per second). Electrodermal activity (EDA), on the other hand, is measured at a rate of 4Hz (i.e., 4 times per second).  Given the high frequency nature of sensor data, this paper assumes the process is measured continuously.

Let~$H_{i,t}^{NX} = H_{i,t}^{N} \otimes H_{i,t}^{X}$ be the $\sigma$-field generated by all past values~$(N_i (s), x_i (s))_{0 \leq s \leq t}$. In this paper, the instantaneous risk of an event at time~$t$ is assumed to depend on the health process, time-in-study, and the event history through a fully parametric conditional hazard function:
\begin{equation}
\label{eq:hazard}
h_i \left( t \Given H_{i,t}^{NX} ; \theta \right) =
\lim_{\delta \to 0} \delta^{-1} \pr \left( N_i(t+\delta) - N_i(t) > 0
  \given H_{i,t}^{NX} \right),
\end{equation}
where~$\theta$ is the parameter vector. For high frequency physiological data, we assume that current risk is log-additive and depends on a linear functional of the health process over some recent window of time and pre-specified features of the counting process; that is,
\begin{equation}
\label{eq:hazardlinear}
h_i \left( t \given  H_{i,t}^{NX} ; \theta \right) =
h_0 (t; \gamma) \exp \left( g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + \int_{t-\Delta}^t x_i (s) \beta(s) ds  \right)
\end{equation}
where~$h_0(t;\gamma)$ is a parametrized baseline hazard function,~$\Delta$ is an unknown window-length, and $g_t( H_{i,t}^N ) \in \mathbb{R}^p$ is a $p$-length feature vector summarizing the event-history and time-in-study information. The final term~$\int_{t-\Delta}^t x_i(s) \beta(s) ds$ reflects the unknown linear functional form of the impact of the time-varying covariate on current risk.

An alternative to~\eqref{eq:hazardlinear} would be to construct features from the sensor data history~$f_t ( H_{i,t}^{X}) \in \mathbb{R}^q$ and incorporated these features in the place of the final term. Our current approach builds linear features of~$H_{i,t}^X$ directly from the integrated history, avoiding the feature construction problem -- a highly nontrivial issue for high frequency time-series data.  The main caveat is the additional parameter~$\Delta$; however, as long as the estimated~$\hat \Delta$ exceeds~$\Delta$, then resulting estimation is unbiased albeit at a loss of efficiency.  Moreover, sensitivity analysis can be performed to determine how choice of $\hat \Delta$ affects inference.  One limitation of the approach presented here is that only fully parametric hazard models may be fit to the data. However, a spline model for the log baseline hazard affords sufficient model flexibility.

\subsection{Measurement-error models with event processes}
\label{section:memproblems}

One potential criticism for~\eqref{eq:hazard} is that the health process may be measured with error. A common mathematical strategy for joint models is to consider an unobservable, latent process~$\eta_i$ such that~$\bfT_i \indep x_i \given \eta_i$, i.e., the two processes are conditionally independent given the latent trajectory. For example, take~$\eta_i$ to be a zero-mean Gaussian process with
\begin{equation}\label{eq:jm}
x_i(t) = \eta_i (t) + \epsilon_i (t),\quad \text{ and } \log h_i (t
\given \eta ) = \log h_0 (t) + g_t \left( H_{i,t}^N \right)^{\prime}
\alpha + \int_{t-\Delta}^t \eta_i (s) \beta (s) ds
\end{equation}
where~$\epsilon_i(t)$ is a white-noise measurement error term. Thus, $\eta_i (t)$ is the ``true and unobserved value of the longitudinal outcome'' \citep[Sec. 2.1, pp.3]{Rizopoulos2010}. The conditional survival function of the $j$th event~$T_j$ given~$\bfx_i$ and all prior events~$\bfT_{-j} := \{ T_1, \ldots, T_{j-1} \}$ is
\[
\pr \left ( T_j > t+s \Given \H_{i,\tau_i}^X, \bfT_{-j} = \bft_{-j}, T_j > t \right) = E \left( - \exp \left( \int_{t}^{t+s} h_i (u \given \eta ) du \right) \Given \bfx_i, \bfT_{-j} = \bft_{-j} \right),
\]
where the expectation is a Gaussian integral, albeit infinite-dimensional.  For most choices of covariance structure, if the white-noise error term is non-zero then the above calculation will show the conditional survival function depends not only on past $x$-values, but also on future $x$-values, i.e.,~$\bfx$ and~$\bfT$ do not satisfy \emph{independent evolution}~\citep{DempseyPMCC2}. This is quite unnatural, as~\eqref{eq:jm} suggests the instantaneous risk of an event at time~$t$ depends on future values of the sensor process. To ensure independent evolution, we set~$\epsilon (t) \equiv 0$ and treat~$\bfx_i$ as a Gaussian process measured without a white-noise error term.
Our procedure will only rely on the Gaussian assumption when~$\bfx$ is not fully observed; see Section~\ref{section:missingdata} for details on handling of missing data.

\subsection{Likelihood calculation}

For the sake of notational simplicity, we leave the dependency of the conditional hazard function on~$H_{i,t}^{NX}$ implicit, and write~$h_i (t; \theta)$. The component of the log-likelihood related to the event process is then given by
\[
  L_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i}
    \log \left( h_i \left( T_{i,j}; \theta
      \right) \right) - H_{i} \left( \tau_i; \theta \right) \right)
\]
where~$H_{i} (\tau_i ; \theta) = \int_{0}^{\tau_i} h_{i} (t; \theta) dt$ is the cumulative hazard function. Solving the associated score equations~$U_n (\theta) = {\bf 0}$ yields the maximum likelihood estimator~$\hat \theta$, where
\[
U_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i} \frac{h^{(1)}_i
    (T_{i,j}; \theta)}{h_i (T_{i,j}; \theta)} - H^{(1)}_{i} (\tau_i;
  \theta) \right),
\]
with $h_i^{(1)} (T_{i,j}; \theta)$ and $H_i^{(1)} (\tau_{i}; \theta)$ are derivatives with respect to $\theta$.

In classical joint models~\citep{Henderson2000, Tsiatis2004}, time-varying covariates~$x_i (t)$ are observed only intermittently at appointment times. In our current setting, maximizing the likelihood is computationally prohibitive since for any~$\theta$ we must compute the cumulative hazard functions~$H_{i} (\tau_i; \theta)$ which requires integration of~$h_i(t;\theta)$ given by~\eqref{eq:hazardlinear} which itself depends on the integral~$\int_{t-\Delta}^t x_i (s) \beta(s) ds$ that is a function of the unknown functional parameter $\beta(\cdot)$.  That is, the risk model now depends on an integrated past history of the time-varying covariate which leads to severe increase in computational complexity.

\subsection{Probabilistic subsampling framework}

To solve the computational challenge we employ a point-process subsampling design to obtain unbiased estimates of the derivative of the cumulative hazards for each subject. The subsampling procedure treats the collected sensor data as a set of \emph{potential observations}. Suppose covariate information is sampled at times drawn from an independent inhomogeneous Poisson point process with known intensity~$\pi_i (t)$. At a subsampled time~$t$, the \emph{windowed covariate history} $\{ x_i (t-s)\}_{0 \leq s \leq \Delta}$ and counting process features~$g_t (H_{i,t}^N)$ are observed. Optimal choice of~$\pi_i (t)$ is beyond the scope of this paper; however, simulation studies have suggested setting the subsampling rate proportional to the hazard function~$h_i (t; \theta)$.

An estimator is design-unbiased if its expectation is equal to that parameter under the probability distribution induced by the sampling design~\citep{Cassel1977}. Let~$\bfD_i \subset [0,t_i]$ denote the random set of subsampled points.  Note, by construction, this random set is distinct from the set of event times with probabilty one, i.e.,~$\pr( \bfT_i \cap \bfD_i = \emptyset) = 1$.  Under subsampling via $\pi_i (t)$, one may compute a Horvitz-Thompson estimator of the derivative of the cumulative hazard $\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in \bfD_i} h^{(1)}_i (u; \theta)/\pi_i (u)$. An alternative design-unbiased estimator of the derivative of the cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in (\bfT_i \cup \bfD_i)}
\frac{ h^{(1)}_i ( u; \theta ) }{ \pi_i (u) + h_i ( u ; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested by~\cite{Waagepetersen2008}.  This estimator depends on the superposition of the event and subsampling processes. Proposition~\ref{prop:optimal} shows the estimator for~$\theta$ associated with using~\eqref{eq:WPest} is the most efficient within a suitable class of estimators for the derivative of the cumulative hazard function (including the Horvitz-Thompson estimator). Therefore, we restrict our attention to~\eqref{eq:WPest} for the remainder of this paper. Letting
\begin{equation}
\label{eq:waage_weights}
w_i (t; \theta) = \frac{\pi_i (t)}{\pi_i (t) + h_i (t ; \theta)},
\end{equation}
the resulting approximate estimating equations can be re-written as
\begin{equation}
\label{eq:approxscore}
\hat{U}_n (\theta) = \sum_{i=1}^n \left[ \sum_{u \in \bfT_i} w_i(u; \theta)
  \frac{h^{(1)} (u; \theta)}{ h ( u; \theta)}  - \sum_{u \in \bfD_i} w_i(u;
  \theta) \frac{h_i^{(1)} (u; \theta)}{ \pi_i (u) } \right].
\end{equation}
Equation~\eqref{eq:approxscore} represents the approximate score functions built via plug-in of the design-unbiased estimator of the derivative of the cumulative hazard given in~\eqref{eq:WPest}.

\begin{rmk}[Connection to design-based estimation]
The approximate score equations given by~\eqref{eq:approxscore} arise in design-based inference of point processes.  Design-based inference is common for spatial point processes~\citep{Waagepetersen2008} where the spatial varying covariate is observed at a random sample of locations. It is common in mobile health where ecological momentary assessments~\citep{Rathbun2012,Rathbun2016} are used to randomly sample individuals at various time-points to assess their emotional state.  In the current setting, we leverage these ideas to form a subsampling protocol that can substantially reduce computationally complexity.  Therefore, the purpose is quite different. Moreover, the dependence of the intensity function on the recent history of sensor values leads to additional complications that must be addressed.
\end{rmk}

\section{Longitudinal functional principal components within
  event-history analysis}

Probabilistic subsampling converts the single sensor stream~$\bfx_i$ into a sequence of functions observed repeatedly at sampled times~$\bfD_i$ and event times~$\bfT_i$ over windows of length~$\Delta$. Such a data structure is commonly referred to as \emph{longitudinal functional data}~\citep{Xiao2013, GoldSmith2015}. Given the large increase in longitudinal functional data in recent years, corresponding analysis has received much recent attention~\citep{Morris2003, MorrisCarroll2006, Baladandayuthapani2008, Di2009, Greven2010, Staicu2010, ChenMuller2012, LiGuan2014}. Here, we combine work by~\cite{Park2018} and~\cite{Goldsmith2011} to construct a computationally efficient penalized functional method for solving the estimation equations~$\hat U_n (\theta)$.

\subsection{Estimation of the windowed covariate history}
\label{sec:margcov}
We start by defining~$X(t,s) = x(t-s)$ to be the sensor measurement~$0 \leq s \leq \Delta$ time units prior to time~$t \in \bfT_i \cup \bfD_i$. The sandwich smoother~\citep{Xiao2013} is used to estimate the mean~$\mu_y(t,s) = \E_y [ X(t,s)]$ where the expectation is indexed by whether $t$ is an event ($y=1$) or subsampled ($y=0$) time respectively. Alternative bivariate smoothers exist, such as the kernel-based local linear smoother~\citep{Hastie2009}, bivariate tensor product splines~\citep{Wood2006}, and the bivariate penalized spline smoother~\citep{MarxEilers2005}. The sandwich smoother was chosen for its computational efficiency and estimation accuracy. We then define~$\tilde X(t,s) = X(t,s) - \hat \mu_y(t,s)$ to be the mean-zero process at each time~$t \in \bfT_i \cup \bfD_i$.

>>>>

\usepackage{xr}
\makeatletter
\newcommand*{\addFileDependency}[1]{% argument=file name and extension
  \typeout{(#1)}% latexmk will find this if $recorder=0 (however, in that case, it will ignore #1 if it is a .aux or .pdf file etc and it exists! if it doesn't exist, it will appear in the list of dependents regardless)
  \@addtofilelist{#1}% if you want it to appear in \listfiles, not really necessary and latexmk doesn't use this
  \IfFileExists{#1}{}{\typeout{No file #1.}}% latexmk will find this message if #1 doesn't exist (yet)
}
\makeatother

\newcommand*{\myexternaldocument}[1]{%
    \externaldocument{#1}%
    \addFileDependency{#1.tex}%
    \addFileDependency{#1.aux}%
}
%%% END HELPER CODE

% put all the external documents here!
\myexternaldocument{fda-recurrent}

@@

\title[Supplement to ``Recurrent event analysis with functional covariates via random subsampling'']{Supplementary Materials to ``Recurrent event analysis in the presence of real-time high frequency data via random subsampling''}
% \author{Walter Dempsey}
% \address {Department of Biostatistics, University of Michigan, 1415 Washington Heights, Ann Arbor, MI 48109, USA}
% \email{wdem@umich.edu}

@@

\begin{abstract}
This is the supplementary materials to the manuscript titled ``Recurrent event analysis in the presence of real-time high frequency data via random subsampling.''  Section~\ref{app:prooflemma31} contains the proofs for Lemmas~\ref{lemma:logistic}, Lemma~~\ref{lemma:simpleasym}.  Section~\ref{app:optimalweights} contains that proof of Proposition~\ref{prop:optimal}.
Section~\ref{app:extrasims} contains additional simulations and case study details.
\end{abstract}

@@

\appendix

\section{Derivation of the design-unbiased score equations}
\label{app:prooflemma31}
